Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-24 _tex/index.tex
Built site for gh-pages

\begin{abstract}
Dichotomous Comparative Judgment
\citep[DCJ,][@Pollitt\_2012b]{Pollitt_2012a} requires judges to evaluate
the relative manifestation of traits between pairs of stimuli, resulting
in a dichotomous outcome indicating which stimulus exhibits the trait
more strongly. Research has demonstrated DCJ's effectiveness and
reliability in various domains
\citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature lacks
a transparent depiction of the DCJ system and the plausible mechanisms
that generate the DCJ data. Particularly, there is no detailed
explanation of how different assessment factors can potentially
influence the observed DCJ data. This study aims to fill this gap by
applying the framework of causal analysis and Directed Acyclic Graphs
{[}DAG; \citet{Pearl_2009}{]}. Using this framework, the study will
construct a scientific model to elucidate the causal assumptions and
mechanisms inherent the system. This model will enable researchers to
draw inferences about causal relationships from DCJ data. Subsequently,
the study will translate this model into a probabilistic statistical
model, aiming to derive statistical estimands for different targets of
inference. The outcomes of this study will inform the planning of DCJ
experiments and hold significance for researchers or analysts involved
in education and assessment procedures who implement the DCJ
methodology.
\end{abstract}

@@

Ultimately, this research aims to extend the law of comparative judgment
initially proposed by \citet{Thurstone_1927} and provide a sound
probabilistic base for the statistical analysis of DCJ data.
Consequently, this research holds significance for researchers and
analysts involved in education and assessment procedures who implement
or design DCJ experiments.

@@

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical representation for analyzing causes and counterfactuals
\citep{Pearl_2009}.

According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
highest level of cognitive abstraction in the ladder of causation,
followed by intervention and association, and form the foundation of
causal inference. Counterfactuals represent scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes resulting from a
cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
abstract nature, counterfactuals enable the development of a
\emph{theory of the world} that explains why specific causes have
specific effects and what occurs in the absence of those causes
\citep{Pearl_et_al_2018}. Counterfactuals achieves this by translating
causal statements to statements about ``what would have happened under
different circumstances.''

@@

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
representations provide a transparent depiction of a system's
complexity, revealing its underlying assumptions and the plausible
causal mechanisms responsible for generating the system's outcome.
Ultimately, DAGs play the important role of effectively conveying the
assumed causal structure of the system, and also determine wich
variables should be controlled for to be able to estimate causal
effects. Consequently, the structural approach will be the primary
method used to address one of the study's research goals.

@@

As previously mentioned, the structural approach employs Directed
Acyclic Graphs (DAGs) for the graphical representation of systems. A
\emph{graph} consists of nodes connected by edges
\citep{Gross_et_al_2018}. In a \emph{directed graph}, edges extend from
a \emph{parent} node to a \emph{child} node, with arrows indicating the
direction of causal influence. In directed \emph{acyclic} graphs, the
direction of causal influences does not loop back on themselves, meaning
the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.

>>>>

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citet{Pollitt_2012b}; \citet{Jones_2015}; \citet{vanDaal_et_al_2016};
\citet{Bartholomew_et_al_2018}; \citet{Lesterhuis_2018};
\citet{Bartholomew_et_al_2020}; \citet{Marshall_et_al_2020};
\citet{Boonen_et_al_2020}{]}. However, despite the method's widespread
use, existing literature lacks a clear explanation of the complexities
and assumptions underpinning the DCJ system, as well as the plausible
mechanisms through which DCJ data are generated. This study addresses
these issues by integrating DCJ into the framework of causal inference.
Specifically, utilizing a structural approach to causal inference, the
study develops a scientific model to clarify the causal assumptions and
mechanisms inherent in the DCJ system. It then translates this model
into a probabilistic statistical framework to estimate statistical
relationships and infer causal connections within the system. This
research builds upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} and provide a robust probabilistic
foundation for the statistical analysis of DCJ data. The findings of
this study offer valuable insights for researchers and analysts involved
in education and assessment procedures who implement or design DCJ
experiments.
\end{abstract}

@@

Ultimately, this research builds upon Thurstone's law of comparative
judgment \citeyearpar{Thurstone_1927} and provide a robust probabilistic
foundation for the statistical analysis of DCJ data. Consequently, the
findings of this study offer valuable insights for researchers and
analysts involved in education and assessment procedures who implement
or design DCJ experiments.

@@

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical notation for analyzing causes and counterfactuals
\citep{Pearl_2009}.

According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
highest level of cognitive abstraction in the ladder of causation,
followed by intervention and association, and form the foundation of
causal inference. Counterfactuals represent scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes resulting from a
cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
abstract nature, counterfactuals enable the development of a
\emph{theory of the world} that explains why specific causes have
specific effects and what occurs in their absence
\citep{Pearl_et_al_2018}. They achieve this by translating causal
statements into counterfactual statements, that is, statements about
``what would have happened in the world under different circumstances.''

@@

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. In this context, DAGs
fulfill the role of a heuristic, effectively conveying the assumed
causal structure of a system. They do not represent detailed statistical
models but allow researchers to deduce which statistical models can
provide valid causal inferences, assuming the causal structure depicted
in the DAG is accurate \citep{McElreath_2020}.

@@

Graph theory is the branch of mathematics focused on the study of
graphs. Graphs are mathematical structures used to model pairwise
relations between objects \citep{Gross_et_al_2018}. While graph theory
covers a wide array of topics, the field of causal inference,
particularly its structural approach, has incorporated some of its
concepts to represent causes and counterfactuals formally and
transparently. A causal graph, or Directed Acyclic Graph (DAG), as its
name suggest, is a directed graph without cycles. A \emph{graph} is a
collection of nodes connected by edges. In a \emph{directed graph},
edges extend from a node to another node, with arrows indicating the
direction of causal influence. In a directed \emph{acyclic} graph, the
direction of causal influences does not loop back on itself, ensuring
that the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.
